A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-17 text/properties.tex
Revised Section 4.

In this section, we prove \AI{use ``show'' because we omit proofs from the paper} basic properties of \LMD: preservation, strong normalization, confluence, and progress.
These theorem shows that the type system of \LMD is strong enough to ensure a typed term of \LMD never go wrong.
\AI{In this setting, a simple type system is already sufficient to guarantee
  ``a well-typed term never goes wrong.''  If the language has
  case analysis that uses dependent types, then
  we could say more.
}

@@

Substitution Lemma in \LMD\ is little more complicated than an ordinary one 
because there are six types of judgment and two types of substitution in \LMD.
Substitution Lemma on Terms states that term substitution $[z \mapsto P]$ preserves
typing, kinding, well-formed kinding, term equivalence, type equivalence, and king equivalence.
Substitution Lemma on Terms states the same for stage substitution $[\beta\mapsto\varepsilon]$.

\begin{lemma}[Substitution Lemma on Terms]
	\begin{flalign*}
		\text{If\ } \G,z:\xi @B \V M:\tau @A \text{\ and\ } \G\V P:\xi @B
		&\text{\ then\ } \G\V M[z \mapsto P]:\tau[z \mapsto P] @A.&\\
		\text{If\ } \G,z:\xi @B \V \tau::K @A \text{\ and\ } \G\V P:\xi @B
		&\text{\ then\ } \G\V \tau[z \mapsto P]::K[z \mapsto P] @A.&\\
		\text{If\ } \G,z:\xi @B \V K\iskind @A \text{\ and\ } \G\V P:\xi @B
		&\text{\ then\ } \G\V K[z \mapsto P] \iskind  @A.&\\
		\text{If\ } \G,z:\xi @B \V M\E N : \tau @A \text{\ and\ } \G\V P:\xi @B
		&\text{\ then\ } \G\V M[z \mapsto P]\E N[z \mapsto P] : \tau[z \mapsto P] @A.&\\
		\text{If\ } \G,z:\xi @B \V \tau\E \sigma : K @A \text{\ and\ } \G\V P:\xi @B
		&\text{\ then\ } \G\V \tau[z \mapsto P]\E \sigma[z \mapsto P] : K[z \mapsto P] @A.&\\
		\text{If\ } \G,z:\xi @B \V K\E J @A \text{\ and\ } \G\V P:\xi @B
		&\text{\ then\ } \G\V K[z \mapsto P]\E J[z \mapsto P] @A.&
	\end{flalign*}
\end{lemma}

\AI{Since the order of variable declarations is significant,
  we cannot assume $z$ is always declared at the end.}

@@

\begin{theorem}[Substitution Lemma on Stages]
	\begin{flalign*}
		\text{If\ } \G \V M:\tau @A
		&\text{\ then\ } \G[\beta \mapsto \varepsilon]\V M[\beta \mapsto \varepsilon]:\tau[\beta \mapsto \varepsilon] @A[\beta \mapsto \varepsilon].&\\
		\text{If\ } \G \V \tau::K @A
		&\text{\ then\ } \G[\beta \mapsto \varepsilon]\V \tau[\beta \mapsto \varepsilon]::K[\beta \mapsto \varepsilon] @A[\beta \mapsto \varepsilon].&\\
		\text{If\ } \G \V K\iskind @A
		&\text{\ then\ } \G[\beta \mapsto \varepsilon]\V K[\beta \mapsto \varepsilon] \iskind @A[\beta \mapsto \varepsilon].&\\
		\text{If\ } \G \V M\E N : \tau @A
		&\text{\ then\ } \G[\beta \mapsto \varepsilon]\V M[\beta \mapsto \varepsilon]\E N[\beta \mapsto \varepsilon] : \tau[\beta \mapsto \varepsilon]  @A[\beta \mapsto \varepsilon].&\\
		\text{If\ } \G \V \tau\E \sigma : K @A
		&\text{\ then\ } \G[\beta \mapsto \varepsilon]\V \tau[\beta \mapsto \varepsilon]\E \sigma[\beta \mapsto \varepsilon] : K[\beta \mapsto \varepsilon] @A[\beta \mapsto \varepsilon].&\\
		\text{If\ } \G \V K\E J @A
		&\text{\ then\ } \G[\beta \mapsto \varepsilon]\V K[\beta \mapsto \varepsilon]\E J[\beta \mapsto \varepsilon] @A[\beta \mapsto \varepsilon].&
	\end{flalign*}
\end{theorem}

@@

The following three Inversion Lemmas are needed to prove main theorems.
As usual~\cite{TAPL}, Inversion Lemmas enable us to infer the types of subterms of a term from the type of the term.
They are composed of three parts.
The first part states that we can infer types of subterms.
The second and third parts state 
that when we have a type equivalence, we can infer the shape of type from the left to the right, and vice versa.
We need both of the second and the third because of the reflexivity of the type equivalence. \AI{Symmetry?  We can omit the third one.}

\begin{lemma}[Inversion Lemma for $\Pi$-types]
	If $\G \V (\lambda x:\sigma.M) : (\Pi x:\sigma'.\tau)@A$ then
	\begin{enumerate}
		\item $\G \V \sigma \E \sigma'@A$ and
		\item $\G ,x:\sigma@A\V M:\tau @A$.
	\end{enumerate}
	\item If $\G \V \rho \E (\Pi x:\sigma.\tau) : K @A$ then $\exists \sigma', \tau', K, J$ such that
	\begin{enumerate}
		\item $\rho = \Pi x:\sigma'.\tau'$,
		\item $\G \V \sigma \E \sigma' : K @A$, and
		\item $\G, x:\sigma@A\V \tau \E \tau' : J @A$.
	\end{enumerate}
	\item If $\G \V (\Pi x:\sigma.\tau) \E \rho : K @A$ then $\exists \sigma', \tau', K, J$ such that
	\begin{enumerate}
		\item $\rho = \Pi x:\sigma'.\tau'$,
		\item $\G \V \sigma \E \sigma' : K @A$, and
		\item $\G, x:\sigma@A\V \tau \E \tau' : J @A$.
	\end{enumerate}
\end{lemma}

\begin{theorem}[Inversion Lemma for $\TW$-types]
	\begin{item}
	      \item If $\G \V \TB_\alpha M : \TW_\alpha \tau @A$ then $\G \V M : \tau @A$.
	      \item If $\G \V \rho \E  \TW_\alpha \tau : K @A$ then $\exists \tau', K, J$ such that
	      \begin{enumerate}
		      \item $\rho = \TW_\alpha \tau'$ and
		      \item $\G @A\V \tau \E \tau' : K @A$.
	      \end{enumerate}
	      \item If $\G \V \TW_\alpha \tau \E \rho : K @A$ then $\exists \tau', K, J$ such that
	      \begin{enumerate}
		      \item $\rho = \TW_\alpha \tau'$ and
		      \item $\G @A\V \tau \E \tau' : K @A$.
	      \end{enumerate}
	\end{item}
\end{theorem}

\begin{theorem}[Inversion Lemma for $\forall$-types]
	\begin{item}
	      \item If $\G \V \Lambda\alpha.M : \forall\alpha.\tau @A$ then $\G \V M : \tau @A$ and $\alpha \notin \FTV(\G) \cup \FV(A)$.
	      \item If $\G \V \rho \E \forall\alpha.\tau : K @A$ then $\exists \tau', K$ such that
	      \begin{enumerate}
		      \item $\rho = \forall\alpha.\tau'$ and
		      \item $\G \V \tau \E \tau' : K @A$.
	      \end{enumerate}
	      \item If $\G \V \forall\alpha.\tau \E \rho : K @A$ then $\exists \tau', K$ such that
	      \begin{enumerate}
		      \item $\rho = \forall\alpha.\tau'$ and
		      \item $\G \V \tau \E \tau' : K @A$.
	      \end{enumerate}
	\end{item}
\end{theorem}

@@

Unique Decomposition ensure that
,for every typed term, we can find just one redex to reduce by the evaluation context or it is a value.
This theorem is important because it guarantee
that the evaluation context decides a redex to reduce deterministically.
Specifically speaking, this theorem guarantee that 
when you write a interpreter using the evaluation context of \LMD,
your interpreter works just as intended.

\begin{theorem}[Unique Decomposition]
	If $x:\tau @\varepsilon \notin \G$ and $\G \V M : \tau @ A$ then either
	\begin{enumerate}
		\item $ M \in V^A$, or
		\item there exist $B, E^A_B$, and $R^B$ such that $M = E^A_B[R^B]$ with $B = \varepsilon$ or $B = \beta$ for some $\beta$.
	\end{enumerate}
\end{theorem}

@@

\begin{theorem}[Progress]
	If $x:\tau @\varepsilon \notin \G$ and $\G \V M : \tau  @ A$ then
	$ M \in V^A $ or $M'$ exists such that $M \longrightarrow_s M'$.
\end{theorem}

@@

\AI{Explain the condition on $\G$.}

>>>>

In this section, we show basic properties of \LMD: preservation, strong normalization, confluence, and progress.

@@

Substitution Lemma in \LMD\ is little more complicated than an ordinary one 
because there are six types of judgment and two types of substitution in \LMD.
Substitution Lemma on Terms states that term substitution $[z \mapsto P]$ preserves
typing, kinding, well-formed kinding, term equivalence, type equivalence, and king equivalence.
Substitution Lemma on Terms states the same for stage substitution $[\beta\mapsto A]$.
\red{2つの代入補題を指してSubstitution Lemmasということが出来るのか?}

In order to write Substitution Lemma briefly, we write $\G\V J@A$ for an arbitary judgment
amongst $\G\V M:\tau@A$, $\G\V \tau::K@A$, $\G\V K \iskind @A$, $\G\V M \E N : \tau @ A$, 
$\G\V \tau \E \sigma @ A$, and $\G \V K \E J @ A$.

\begin{lemma}[Substitution Lemma on Terms]
	If $\G, z:\xi@B, \D \V J @ A$ and $\G\V P:\xi @B$ then $\G, \D[z \mapsto P] \V J[z \mapsto P] @ A$.
\end{lemma}

\red{TODO: Fix the proof.}
\red{$\G\V J@A$における代入を定義する必要があるか?}

@@

\begin{lemma}[Substitution Lemma on Stages]
	If $\G \V J @ A$ then $\G[\beta\mapsto B] \V J[\beta\mapsto B] @ A[\beta\mapsto B]$.
\end{lemma}

@@

The following three Inversion Lemmas are needed to prove main theorems.
As usual~\cite{TAPL}, Inversion Lemmas enable us to infer the types of subterms of a term from the type of the term.

\begin{lemma}[Inversion Lemma for $\Pi$-types]
	If $\G \V (\lambda x:\sigma.M) : (\Pi x:\sigma'.\tau)@A$
	then $\G \V \sigma \E \sigma'@A$ and $\G ,x:\sigma@A\V M:\tau @A$.
\end{lemma}

\begin{proof}
	First, we generalize this theorem by adding statements about $\Pi$-types in equivalence relation in order to use induction.
	After that, we can prove with straightforward induction on derivations.
\end{proof}

\begin{lemma}[Inversion Lemma for $\TW$-types]
	If $\G \V \TB_\alpha M : \TW_\alpha \tau @A$ then $\G \V M : \tau @A$.
\end{lemma}

\begin{proof}
	Prove as with Inversion Lemma for $\Pi$-types.
\end{proof}

\begin{lemma}[Inversion Lemma for $\forall$-types]
	If $\G \V \Lambda\alpha.M : \forall\alpha.\tau @A$ then $\G \V M : \tau @A$ and $\alpha \notin \FTV(\G) \cup \FV(A)$.
\end{lemma}

\begin{proof}
	Prove as with Inversion Lemma for $\Pi$-types.
\end{proof}

@@

Unique Decomposition ensure that,
for every typed term, we can find just one redex to reduce by the evaluation context or it is a value.
This theorem is important because it guarantee
that the evaluation context decides a redex to reduce deterministically.
Specifically speaking, this theorem guarantee that 
when you write a interpreter using the evaluation context of \LMD,
your interpreter works just as intended.
Although, in ordinary calculi, there is a condition that $\G$ is $\emptyset$,
the condition is relaxed because variables at non-$\varepsilon$ stages are values in \LMD.

\begin{theorem}[Unique Decomposition]
	If $\G$ does not have any variable declared at stage $\varepsilon$ 
	and $\G \V M : \tau @ A$ then either
	\begin{enumerate}
		\item $ M \in V^A$, or
		\item there exist $B, E^A_B$, and $R^B$ such that $M = E^A_B[R^B]$ with $B = \varepsilon$ or $B = \beta$ for some $\beta$.
	\end{enumerate}
\end{theorem}

@@

\begin{theorem}[Progress]
	If $\G$ does not have any variable declared at stage $\varepsilon$ and $\G \V M : \tau  @ A$ then
	$ M \in V^A $ or $M'$ exists such that $M \longrightarrow_s M'$.
\end{theorem}

@@
